---
title: "Psychometric Evaluation of the PHQ-9 in U.S. Adults Using NHANES 2017–March 2020"
subtitle: "Reliability, Dimensionality, Functional Difficulty, and Clinical Interpretation"
author: "Shane Hodson"
date: "22 July 2026"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
    fig-cap-location: bottom
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(dplyr)
library(readr)
library(here)
library(knitr)

display_audit_table <- function(file_name, caption, digits = 2) {

  file_path <- here("tables", file_name)

  if (file.exists(file_path)) {

    read_csv(
      file_path,
      show_col_types = FALSE
    ) |>
      kable(
        caption = caption,
        digits = digits
      )

  } else {

    cat(
      paste0(
        "**Output not yet generated:** `",
        file_name,
        "`"
      )
    )
  }
}

read_stage2a_table <- function(file_name) {

  file_path <- here("tables", file_name)

  if (!file.exists(file_path)) {
    stop(
      paste0(
        "Required Stage 2A table not found: ",
        file_path
      )
    )
  }

  read_csv(
    file_path,
    show_col_types = FALSE
  )
}

format_percent_ci <- function(
  estimate,
  lower,
  upper,
  digits = 2
) {

  paste0(
    formatC(
      estimate,
      format = "f",
      digits = digits
    ),
    "% (",
    formatC(
      lower,
      format = "f",
      digits = digits
    ),
    "–",
    formatC(
      upper,
      format = "f",
      digits = digits
    ),
    "%)"
  )
}

display_figure <- function(file_name) {

  figure_path <- here(
    "figures",
    file_name
  )

  if (file.exists(figure_path)) {

    include_graphics(
      figure_path
    )

  } else {

    cat(
      paste0(
        "**Figure not yet generated:** `",
        file_name,
        "`"
      )
    )
  }
}

# Stage 2B reliability and item-analysis outputs

item_floor_sparsity_summary <- readr::read_csv(
  here::here(
    "tables",
    "phq9_item_floor_sparsity_summary.csv"
  ),
  show_col_types = FALSE
)

corrected_item_total_correlations <- readr::read_csv(
  here::here(
    "tables",
    "phq9_corrected_item_total_correlations.csv"
  ),
  show_col_types = FALSE
)

polychoric_matrix <- readr::read_csv(
  here::here(
    "tables",
    "phq9_polychoric_matrix.csv"
  ),
  show_col_types = FALSE
)

polychoric_matrix_summary <- readr::read_csv(
  here::here(
    "tables",
    "phq9_polychoric_matrix_summary.csv"
  ),
  show_col_types = FALSE
)

ordinal_alpha_summary <- readr::read_csv(
  here::here(
    "tables",
    "phq9_ordinal_alpha_summary.csv"
  ),
  show_col_types = FALSE
)

ordinal_alpha_if_item_removed <- readr::read_csv(
  here::here(
    "tables",
    "phq9_ordinal_alpha_if_item_removed.csv"
  ),
  show_col_types = FALSE
)

omega_total_summary <- readr::read_csv(
  here::here(
    "tables",
    "phq9_omega_total_summary.csv"
  ),
  show_col_types = FALSE
)

omega_if_item_removed <- readr::read_csv(
  here::here(
    "tables",
    "phq9_omega_if_item_removed.csv"
  ),
  show_col_types = FALSE
)

omega_uniqueness_summary <- readr::read_csv(
  here::here(
    "tables",
    "phq9_omega_uniqueness_summary.csv"
  ),
  show_col_types = FALSE
)

reliability_bootstrap_ci <- readr::read_csv(
  here::here(
    "tables",
    "phq9_reliability_bootstrap_ci.csv"
  ),
  show_col_types = FALSE
)

reliability_bootstrap_diagnostics <- readr::read_csv(
  here::here(
    "tables",
    "phq9_reliability_bootstrap_diagnostics.csv"
  ),
  show_col_types = FALSE
)
```

# Project status

This report documents the completed Stage 1 data audit, Stage 2A weighted descriptive analyses, and Stage 2B reliability and item analysis.

Stage 1 includes:

- source-file, linkage and response-code audits;
- sample-definition and missingness checks;
- evaluation of the `DPQ100` routing pattern;
- validation of the required NHANES survey-design variables.

Stage 2A includes:

- construction and validation of the NHANES survey-design object;
- weighted PHQ-9 item-response proportions;
- weighted PHQ-9 total-score descriptives;
- weighted symptom-severity band estimates;
- weighted `DPQ100` descriptives among routed respondents;
- an extended functional-difficulty distribution retaining structural skips and missing responses as separate categories;
- exported descriptive tables and figures.

Stage 2B includes:

- assessment of item floor effects and response-category sparsity;
- corrected item-total correlations;
- estimation and validation of the polychoric correlation matrix;
- ordinal coefficient alpha;
- McDonald’s omega total;
- reliability estimates after removing individual items;
- nonparametric bootstrap confidence intervals based on 2,000 participant-level resamples;
- exported reliability tables and figures.

Factor analysis, confirmatory modelling, measurement invariance, functional-difficulty regression modelling, diagnostic interpretation and substantive clinical interpretation have not begun.

Stage 2B reliability estimates are interpreted as evidence about internal consistency in the complete-PHQ-9 analytic sample. They are not treated as evidence of validity or unidimensionality.

# Dataset

The project uses the combined NHANES 2017–March 2020 pre-pandemic release.

The primary files are:

- `P_DPQ.xpt`: Mental Health—Depression Screener
- `P_DEMO.xpt`: demographic and survey-design variables

The files are linked using the NHANES respondent identifier `SEQN`.

Raw NHANES files are downloaded programmatically and are not redistributed through the repository.

# Stage 1: Data audit

## Source-file audit

```{r}
display_audit_table(
  "data_file_audit.csv",
  "Initial NHANES source-file audit"
)
```

## SEQN linkage audit

```{r}
display_audit_table(
  "seqn_linkage_audit.csv",
  "Linkage audit using SEQN"
)
```

All 8,965 depression-screener records were successfully matched to the demographic file.

Adults appearing in the demographic file but not in the depression-screener file are retained in the sample-flow documentation but are not added to the depression analysis sample.

## PHQ-9 variable dictionary

```{r}
phq9_dictionary_table <- read_stage2a_table(
  "phq9_variable_dictionary.csv"
) |>
  transmute(
    Variable = variable,
    `Item number` = item_number,
    `Symptom content` = content,
    Role = role,
    `Included in PHQ-9 total` = if_else(
      variable == "DPQ100",
      "No",
      "Yes"
    )
  )

kable(
  phq9_dictionary_table,
  caption = "PHQ-9 and functional-difficulty variable dictionary",
  align = c(
    "l",
    "r",
    "l",
    "l",
    "l"
  )
)
```

## Response-code audit

```{r}
display_audit_table(
  "response_code_audit.csv",
  "Observed questionnaire response codes"
)
```

The audit distinguishes valid responses from refusal, “don’t know” and system-missing values.

No unexpected questionnaire response codes were identified.

## Missingness audit

```{r}
display_audit_table(
  "missingness_audit.csv",
  "Item-level missingness audit"
)
```

## DPQ100 routing audit

```{r}
display_audit_table(
  "dpq100_skip_audit.csv",
  "DPQ100 status by PHQ-9 response pattern"
)
```

All participants with valid zero responses to all nine PHQ-9 items had a system-missing `DPQ100` response.

Among participants endorsing at least one PHQ-9 symptom, nearly all had a valid `DPQ100` response.

This pattern is consistent with structural questionnaire routing among participants reporting no PHQ-9 symptoms.

Structurally skipped `DPQ100` responses are not recoded as “not difficult at all.”

## Initial sample flow

```{r}
display_audit_table(
  "sample_flow.csv",
  "Initial analytic sample flow"
)
```

## Survey-design variable audit

```{r}
display_audit_table(
  "survey_design_audit.csv",
  "NHANES survey-design variable audit"
)
```

The combined-cycle examination weight, masked variance stratum and masked variance primary sampling unit variables were present for all 8,965 linked depression-screener records.

## Cleaned questionnaire-code check

```{r}
display_audit_table(
  "clean_code_check.csv",
  "Questionnaire values after recoding non-substantive responses"
)
```

Refusal, “don’t know” and system-missing responses were treated as missing values.

After cleaning, each PHQ-9 symptom item contained only valid values from 0 to 3 or missing values.

# Stage 2A: Weighted descriptive analyses

## Analytic samples

The linked depression-screener dataset contains 8,965 adults.

The analysis-specific samples are:

- **8,276 participants** with valid responses to all nine PHQ-9 items;
- **5,522 participants** with complete PHQ-9 data and at least one endorsed symptom;
- **5,517 participants** with complete PHQ-9 data, at least one endorsed symptom and valid `DPQ100`;
- **2,754 participants** with zero on all nine PHQ-9 items and structurally skipped `DPQ100`;
- **5 participants** with complete PHQ-9 data, at least one endorsed symptom and missing `DPQ100`.

The complete-PHQ-9 sample is used for total-score and symptom-severity descriptives.

Item-response estimates use all available valid responses for each individual item.

The primary `DPQ100` distribution is restricted to routed respondents with valid responses.

## Survey design

The survey-design object was created from the full linked depression-screener dataset using:

- examination weight: `WTMECPRP`;
- masked variance stratum: `SDMVSTRA`;
- masked variance primary sampling unit: `SDMVPSU`;
- nested PSU identifiers within strata.

The design contains 49 nested PSUs across 24 strata.

No observed stratum contained fewer than two PSUs.

Analysis-specific samples were defined as survey domains after the full survey-design object had been created.

Weighted means and item-response proportions were estimated using `svymean()`. Item-response confidence intervals were calculated from the design-based standard errors.

Weighted variances were estimated using `svyvar()`.

Proportions for individual total scores, severity bands and `DPQ100` categories were estimated using `svyciprop()` with logit-transformed 95% confidence intervals.

## Weighted PHQ-9 item-response proportions

Item-specific valid sample sizes ranged from 8,295 to 8,299.

The weighted response proportions sum to 100% within rounding error for each item.

```{r}
item_response_table <- read_stage2a_table(
  "phq9_item_response_proportions.csv"
) |>
  transmute(
    Item = item,
    `Symptom content` = symptom_content,
    Response = response_label,
    `Unweighted n` = unweighted_n_response,
    `Weighted estimate (95% CI)` = format_percent_ci(
      weighted_percent,
      ci_lower_percent,
      ci_upper_percent
    )
  )

kable(
  item_response_table,
  caption = paste(
    "Weighted PHQ-9 item-response proportions.",
    "Item-specific estimates use all available valid responses."
  ),
  align = c(
    "l",
    "l",
    "l",
    "r",
    "r"
  )
)
```

```{r}
#| fig-alt: "Stacked horizontal bar chart showing weighted response proportions for each of the nine PHQ-9 items."
#| out-width: "100%"

display_figure(
  "phq9_item_response_proportions.png"
)
```

## Weighted PHQ-9 total-score distribution

The complete-PHQ-9 sample contains 8,276 participants.

Observed PHQ-9 total scores range from 0 to 27.

```{r}
total_summary_raw <- read_stage2a_table(
  "phq9_total_summary.csv"
)

total_summary_table <- tibble(
  Measure = c(
    "Unweighted sample size",
    "Weighted mean",
    "Standard error of weighted mean",
    "95% confidence interval for weighted mean",
    "Weighted standard deviation",
    "Observed minimum",
    "Observed maximum"
  ),
  Estimate = c(
    format(
      total_summary_raw$unweighted_n,
      big.mark = ","
    ),
    formatC(
      total_summary_raw$weighted_mean,
      format = "f",
      digits = 2
    ),
    formatC(
      total_summary_raw$standard_error,
      format = "f",
      digits = 2
    ),
    paste0(
      formatC(
        total_summary_raw$ci_lower,
        format = "f",
        digits = 2
      ),
      "–",
      formatC(
        total_summary_raw$ci_upper,
        format = "f",
        digits = 2
      )
    ),
    formatC(
      total_summary_raw$weighted_standard_deviation,
      format = "f",
      digits = 2
    ),
    total_summary_raw$observed_minimum,
    total_summary_raw$observed_maximum
  )
)

kable(
  total_summary_table,
  caption = "Weighted PHQ-9 total-score summary.",
  align = c(
    "l",
    "r"
  )
)
```

The weighted mean PHQ-9 total score was 3.18, with a 95% confidence interval from 3.06 to 3.31.

The weighted standard deviation was 4.09.

```{r}
total_distribution_table <- read_stage2a_table(
  "phq9_total_distribution.csv"
) |>
  transmute(
    `PHQ-9 total` = phq9_total,
    `Unweighted n` = unweighted_n,
    `Weighted estimate (95% CI)` = format_percent_ci(
      weighted_percent,
      ci_lower_percent,
      ci_upper_percent
    )
  )

kable(
  total_distribution_table,
  caption = paste(
    "Weighted distribution of PHQ-9 total scores",
    "in the complete-PHQ-9 sample."
  ),
  align = c(
    "r",
    "r",
    "r"
  )
)
```

```{r}
#| fig-alt: "Bar chart showing the weighted distribution of PHQ-9 total scores from 0 to 27."
#| out-width: "100%"

display_figure(
  "phq9_total_score_distribution.png"
)
```

## Weighted symptom-severity bands

The conventional score bands used in this descriptive analysis are:

- minimal: 0–4;
- mild: 5–9;
- moderate: 10–14;
- moderately severe: 15–19;
- severe: 20–27.

These categories describe score ranges and are not treated as psychiatric diagnoses.

```{r}
severity_table <- read_stage2a_table(
  "phq9_severity_distribution.csv"
) |>
  transmute(
    `Severity band` = severity_band,
    `Unweighted n` = unweighted_n,
    `Weighted estimate (95% CI)` = format_percent_ci(
      weighted_percent,
      ci_lower_percent,
      ci_upper_percent
    )
  )

kable(
  severity_table,
  caption = paste(
    "Weighted PHQ-9 symptom-severity bands",
    "in the complete-PHQ-9 sample."
  ),
  align = c(
    "l",
    "r",
    "r"
  )
)
```

The weighted distribution was 74.92% minimal, 16.54% mild, 5.95% moderate, 1.81% moderately severe and 0.78% severe.

```{r}
#| fig-alt: "Horizontal bar chart showing weighted PHQ-9 symptom-severity band percentages with 95% confidence intervals."
#| out-width: "100%"

display_figure(
  "phq9_severity_distribution.png"
)
```

## DPQ100 among routed respondents

The primary functional-difficulty descriptive sample contains 5,517 participants who:

- had complete PHQ-9 data;
- endorsed at least one PHQ-9 symptom;
- provided a valid `DPQ100` response.

This section reports the weighted response distribution only.

It does not model the relationship between PHQ-9 scores and functional difficulty.

```{r}
dpq100_routed_table <- read_stage2a_table(
  "dpq100_routed_distribution.csv"
) |>
  transmute(
    Response = response_label,
    `Unweighted n` = unweighted_n,
    `Weighted estimate (95% CI)` = format_percent_ci(
      weighted_percent,
      ci_lower_percent,
      ci_upper_percent
    )
  )

kable(
  dpq100_routed_table,
  caption = paste(
    "Weighted DPQ100 distribution among routed respondents",
    "with valid functional-difficulty data."
  ),
  align = c(
    "l",
    "r",
    "r"
  )
)
```

Among routed respondents, 74.64% reported “not difficult at all,” 20.79% reported “somewhat difficult,” 3.40% reported “very difficult,” and 1.17% reported “extremely difficult.”

```{r}
#| fig-alt: "Horizontal bar chart showing weighted DPQ100 response percentages among routed respondents with 95% confidence intervals."
#| out-width: "100%"

display_figure(
  "dpq100_routed_distribution.png"
)
```

## Extended functional-difficulty categories

The extended table uses the complete-PHQ-9 sample of 8,276 participants.

Participants with zero on all nine PHQ-9 items are retained in a separate category:

> No PHQ-9 symptoms / DPQ100 structurally skipped

The five symptom-positive participants without a valid `DPQ100` response are also shown separately.

Neither group is assigned to an observed functional-difficulty response category.

```{r}
dpq100_extended_table <- read_stage2a_table(
  "dpq100_extended_distribution.csv"
) |>
  transmute(
    Category = category,
    `Unweighted n` = unweighted_n,
    `Weighted estimate (95% CI)` = format_percent_ci(
      weighted_percent,
      ci_lower_percent,
      ci_upper_percent
    )
  )

kable(
  dpq100_extended_table,
  caption = paste(
    "Extended weighted functional-difficulty distribution",
    "in the complete-PHQ-9 sample."
  ),
  align = c(
    "l",
    "r",
    "r"
  )
)
```

```{r}
#| fig-alt: "Horizontal bar chart showing extended weighted functional-difficulty categories, including structural skips and missing DPQ100 responses."
#| out-width: "100%"

display_figure(
  "dpq100_extended_distribution.png"
)
```

# Stage 2A validation checks

The following checks were completed:

- the linked depression-screener sample contains 8,965 records;
- the complete-PHQ-9 sample contains 8,276 records;
- the symptom-positive complete-PHQ-9 sample contains 5,522 records;
- the routed valid-`DPQ100` sample contains 5,517 records;
- the structural-skip group contains 2,754 records;
- the weighted response proportions sum to 100% within rounding error for each PHQ-9 item;
- the score-distribution counts sum to 8,276;
- the weighted score-distribution proportions sum to 100% within rounding error;
- the severity-band counts sum to 8,276;
- the weighted severity-band proportions sum to 100% within rounding error;
- the routed `DPQ100` counts sum to 5,517;
- the weighted routed-`DPQ100` proportions sum to 100% within rounding error;
- the extended functional-difficulty counts sum to 8,276;
- the weighted extended-category proportions sum to 100% within rounding error;
- no symptom-free participant had a valid `DPQ100` response;
- five symptom-positive participants had missing `DPQ100` responses and remain classified as missing.

The completed survey-design object is saved locally as:

```text
data/processed/nhanes_design_stage2a.rds
```

The Stage 2A working dataset is saved locally as:

```text
data/processed/phq9_stage2a_data.rds
```

These processed files are excluded from version control.

# Stage 2B: Reliability and item analysis

## Analytic sample

Stage 2B used the complete-PHQ-9 analytic sample of 8,276 participants.

All nine PHQ-9 items had valid scores from 0 to 3 for every participant in this sample. No item-level missing values were present in the Stage 2B item matrix.

The reliability analyses were sample-based psychometric analyses. Survey weights, strata and primary sampling units were not incorporated into the reliability coefficients. The resulting estimates therefore describe internal consistency in the complete-PHQ-9 analytic sample rather than nationally representative reliability parameters.

## Methods

### Item floor effects and response-category sparsity

For each PHQ-9 item, the unweighted frequency and percentage of responses in categories 0, 1, 2 and 3 were calculated.

The percentage responding 0 was used to describe the item floor effect. The percentage responding 3 was also reported.

Response categories containing fewer than 1% of valid responses were flagged as sparse. This threshold was used as a descriptive screening rule only. It was not used as an automatic basis for collapsing response categories or removing an item.

### Corrected item-total correlations

Corrected item-total correlations were calculated as Pearson correlations between each PHQ-9 item and the sum of the remaining eight items.

The focal item was excluded from its comparison score to avoid part–whole inflation.

Two-sided 95% confidence intervals were calculated using Fisher’s $z$ transformation, as implemented by `cor.test()` for Pearson correlations. The manually calculated estimates were validated against the `r.drop` output from `psych::alpha()`.

### Polychoric correlations

A polychoric correlation matrix was estimated because the PHQ-9 items are ordered categorical variables with four response categories.

The matrix was estimated using `psych::polychoric()` with:

- a continuity correction of 0.5 for empty bivariate response cells;
- global thresholds;
- automatic matrix smoothing disabled.

The raw matrix was checked for:

- finite values;
- symmetry;
- unit diagonal values;
- positive definiteness.

Matrix smoothing would have been applied only if the raw matrix was not positive definite. The raw matrix was positive definite and did not require smoothing.

The polychoric correlations are reported as item-association estimates. They are not interpreted as evidence that the PHQ-9 is unidimensional.

### Ordinal coefficient alpha

Ordinal coefficient alpha was calculated from the polychoric correlation matrix.

For $k$ items and mean inter-item polychoric correlation $\bar{r}$, ordinal alpha was calculated as:

$$
\alpha =
\frac{k\bar{r}}
{1 + (k - 1)\bar{r}}
$$

The result was validated against the standardised alpha returned by `psych::alpha()` when applied to the same polychoric matrix.

Ordinal alpha was also recalculated after removing each item individually. These item-removal estimates were treated as diagnostic information rather than automatic item-deletion criteria.

### McDonald’s omega total

McDonald’s omega total was estimated from a single-factor congeneric reliability model fitted to the polychoric correlation matrix using minimum-residual extraction.

The one-factor model was used only to obtain the uniqueness estimates required for omega total. Factor loadings, model fit and alternative dimensional structures were not evaluated during Stage 2B.

For the unit-weighted total score, omega total was calculated as:

$$
\omega_t =
1 -
\frac{\sum_i \theta_i}
{\mathrm{Var}(X)}
$$

where $\theta_i$ denotes the estimated uniqueness of item $i$, and $\mathrm{Var}(X)$ is the observed variance of the standardised unit-weighted total score, calculated as the sum of all elements in the polychoric correlation matrix.

Omega total was also recalculated after removing each item individually. These estimates were treated as diagnostic information and not as automatic item-deletion criteria.

### Bootstrap uncertainty intervals

Uncertainty intervals for ordinal alpha and omega total were estimated using 2,000 ordinary nonparametric participant-level bootstrap resamples.

For each resample:

1. participants were sampled with replacement;
2. the polychoric correlation matrix was re-estimated;
3. positive definiteness was checked;
4. ordinal alpha was recalculated;
5. omega total was recalculated.

The reported 95% confidence intervals are percentile bootstrap intervals.

These are participant-level bootstrap intervals for the complete analytic sample. They are not survey-design-based confidence intervals and should not be interpreted as nationally representative reliability intervals.

## Results

### Item floor effects and response-category sparsity

```{r}
item_floor_table <- item_floor_sparsity_summary |>
  transmute(
    Item = variable,
    `Symptom content` = item,
    `Response 0 (%)` = formatC(
      floor_percentage,
      format = "f",
      digits = 2
    ),
    `Response 3 (%)` = formatC(
      ceiling_percentage,
      format = "f",
      digits = 2
    ),
    `Smallest category (%)` = formatC(
      smallest_category_percentage,
      format = "f",
      digits = 2
    ),
    `Empty categories` = number_empty_categories,
    `Categories below 1%` =
      sparse_response_categories_lt_1pct
  )

kable(
  item_floor_table,
  caption = paste(
    "PHQ-9 item floor effects and response-category sparsity",
    "in the complete-PHQ-9 analytic sample."
  ),
  align = c(
    "l",
    "l",
    "r",
    "r",
    "r",
    "r",
    "l"
  )
)
```

No PHQ-9 item had an empty response category.

The percentage responding 0 ranged from 50.19% for tiredness or low energy (`DPQ040`) to 96.15% for thoughts of death or self-harm (`DPQ090`).

`DPQ090` had the strongest floor effect. Its response categories 2 and 3 each contained fewer than 1% of participants. No response category for any other item fell below the predefined 1% screening threshold.

The sparse upper categories for `DPQ090` were retained. They were not collapsed, and the item was not removed.

### Corrected item-total correlations

```{r}
corrected_item_total_table <-
  corrected_item_total_correlations |>
  transmute(
    Item = variable,
    `Symptom content` = item,
    `Corrected item-total r` = formatC(
      corrected_item_total_r,
      format = "f",
      digits = 3
    ),
    `95% CI` = paste0(
      formatC(
        confidence_interval_lower,
        format = "f",
        digits = 3
      ),
      "–",
      formatC(
        confidence_interval_upper,
        format = "f",
        digits = 3
      )
    )
  )

kable(
  corrected_item_total_table,
  caption = paste(
    "Corrected PHQ-9 item-total correlations.",
    "Each item was correlated with the sum of the remaining eight items."
  ),
  align = c(
    "l",
    "l",
    "r",
    "r"
  )
)
```

All nine corrected item-total correlations were positive.

The correlations ranged from 0.394 for `DPQ090` to 0.682 for depressed mood or hopelessness (`DPQ020`).

The lower correlation for `DPQ090` is considered alongside its strong floor effect and sparse upper response categories. It is not treated as sufficient evidence for deleting the item.

```{r}
#| fig-alt: "Point and interval plot showing corrected item-total correlations for all nine PHQ-9 items."
#| out-width: "100%"

display_figure(
  "phq9_corrected_item_total_correlations.png"
)
```

### Polychoric correlation matrix

```{r}
polychoric_matrix_table <- polychoric_matrix |>
  rename(
    Item = variable
  ) |>
  mutate(
    across(
      -Item,
      ~ round(.x, 3)
    )
  )

kable(
  polychoric_matrix_table,
  caption = paste(
    "Polychoric correlation matrix for the nine PHQ-9 items",
    "in the complete-PHQ-9 analytic sample."
  ),
  align = c(
    "l",
    rep("r", 9)
  )
)
```

The polychoric matrix was finite, symmetric and positive definite. Its minimum eigenvalue was 0.204, and no smoothing was required.

All off-diagonal correlations were positive. They ranged from 0.464 to 0.769, with a mean off-diagonal correlation of 0.570.

The strongest estimated association was between depressed mood or hopelessness (`DPQ020`) and negative self-evaluation (`DPQ060`), $r = .769$.

The weakest estimated association was between appetite disturbance (`DPQ050`) and thoughts of death or self-harm (`DPQ090`), $r = .464$.

These correlations describe relationships among the ordered item responses. Their magnitude is not treated as proof that a one-factor model is adequate.

```{r}
#| fig-alt: "Heatmap showing the polychoric correlations among the nine PHQ-9 items."
#| out-width: "90%"

display_figure(
  "phq9_polychoric_correlation_heatmap.png"
)
```

### Ordinal alpha and omega total

```{r}
reliability_coefficients_table <- reliability_bootstrap_ci |>
  transmute(
    Coefficient = coefficient,
    Estimate = formatC(
      point_estimate,
      format = "f",
      digits = 3
    ),
    `Bootstrap standard error` = formatC(
      bootstrap_standard_error,
      format = "f",
      digits = 4
    ),
    `95% bootstrap CI` = paste0(
      formatC(
        confidence_interval_lower,
        format = "f",
        digits = 3
      ),
      "–",
      formatC(
        confidence_interval_upper,
        format = "f",
        digits = 3
      )
    ),
    `Successful resamples` = format(
      successful_replicates,
      big.mark = ","
    )
  )

kable(
  reliability_coefficients_table,
  caption = paste(
    "Ordinal internal-consistency estimates with",
    "95% nonparametric percentile bootstrap intervals."
  ),
  align = c(
    "l",
    "r",
    "r",
    "r",
    "r"
  )
)
```

Ordinal coefficient alpha was `r formatC(reliability_bootstrap_ci$point_estimate[reliability_bootstrap_ci$coefficient == "Ordinal coefficient alpha"], format = "f", digits = 3)`, with a 95% percentile bootstrap confidence interval from `r formatC(reliability_bootstrap_ci$confidence_interval_lower[reliability_bootstrap_ci$coefficient == "Ordinal coefficient alpha"], format = "f", digits = 3)` to `r formatC(reliability_bootstrap_ci$confidence_interval_upper[reliability_bootstrap_ci$coefficient == "Ordinal coefficient alpha"], format = "f", digits = 3)`.

McDonald’s omega total was `r formatC(reliability_bootstrap_ci$point_estimate[reliability_bootstrap_ci$coefficient == "McDonald's omega total"], format = "f", digits = 3)`, with a 95% percentile bootstrap confidence interval from `r formatC(reliability_bootstrap_ci$confidence_interval_lower[reliability_bootstrap_ci$coefficient == "McDonald's omega total"], format = "f", digits = 3)` to `r formatC(reliability_bootstrap_ci$confidence_interval_upper[reliability_bootstrap_ci$coefficient == "McDonald's omega total"], format = "f", digits = 3)`.

All nine estimated uniqueness values used in the omega-total calculation were finite, non-negative and no greater than 1. They ranged from `r formatC(omega_uniqueness_summary$minimum_uniqueness, format = "f", digits = 3)` to `r formatC(omega_uniqueness_summary$maximum_uniqueness, format = "f", digits = 3)`.

Because omega total was estimated from a one-factor congeneric model, its interpretation will be revisited after the Stage 3 dimensionality analysis.

The bootstrap means were close to the point estimates, and estimated bootstrap bias was negligible.

Ordinal alpha and omega total were high in the complete analytic sample. They provide evidence about internal consistency but do not establish that the PHQ-9 is valid, diagnostically accurate or adequately represented by one dimension.

```{r}
#| fig-alt: "Point and interval plot showing ordinal coefficient alpha and McDonald's omega total with bootstrap confidence intervals."
#| out-width: "90%"

display_figure(
  "phq9_reliability_coefficients.png"
)
```

### Reliability if an item was removed

```{r}
reliability_if_removed_table <-
  ordinal_alpha_if_item_removed |>
  select(
    variable,
    item,
    ordinal_alpha_if_removed,
    ordinal_alpha_change
  ) |>
  left_join(
    omega_if_item_removed |>
      select(
        variable,
        omega_total_if_removed,
        omega_total_change
      ),
    by = "variable"
  ) |>
  transmute(
    Item = variable,
    `Symptom content` = item,
    `Ordinal alpha if removed` = formatC(
      ordinal_alpha_if_removed,
      format = "f",
      digits = 3
    ),
    `Change in alpha` = formatC(
      ordinal_alpha_change,
      format = "f",
      digits = 4
    ),
    `Omega total if removed` = formatC(
      omega_total_if_removed,
      format = "f",
      digits = 3
    ),
    `Change in omega` = formatC(
      omega_total_change,
      format = "f",
      digits = 4
    )
  )

kable(
  reliability_if_removed_table,
  caption = paste(
    "Ordinal alpha and omega total after removing",
    "each PHQ-9 item individually."
  ),
  align = c(
    "l",
    "l",
    "r",
    "r",
    "r",
    "r"
  )
)
```

Removing any individual item lowered both ordinal alpha and omega total.

The largest decreases occurred when `DPQ020` was removed. Ordinal alpha decreased by 0.0160 and omega total decreased by 0.0164.

The item-removal results do not support improving internal consistency by deleting any single PHQ-9 item. They are not interpreted as evidence that every item is interchangeable or that item content should be judged only by its contribution to a reliability coefficient.

### Bootstrap diagnostics

```{r}
bootstrap_diagnostics_table <- tibble(
  Measure = c(
    "Requested bootstrap resamples",
    "Successful bootstrap resamples",
    "Failed bootstrap resamples",
    "Resampled matrices requiring smoothing",
    "Resamples producing computational warnings",
    "Minimum raw matrix eigenvalue across resamples",
    "Elapsed computation time"
  ),
  Result = c(
    format(
      reliability_bootstrap_diagnostics$
        requested_replicates,
      big.mark = ","
    ),
    format(
      reliability_bootstrap_diagnostics$
        successful_replicates,
      big.mark = ","
    ),
    reliability_bootstrap_diagnostics$
      failed_replicates,
    reliability_bootstrap_diagnostics$
      smoothed_matrices,
    reliability_bootstrap_diagnostics$
      replicates_with_warnings,
    formatC(
      reliability_bootstrap_diagnostics$
        minimum_raw_eigenvalue,
      format = "f",
      digits = 3
    ),
    paste0(
      formatC(
        reliability_bootstrap_diagnostics$
          elapsed_minutes,
        format = "f",
        digits = 2
      ),
      " minutes"
    )
  )
)

kable(
  bootstrap_diagnostics_table,
  caption = "Stage 2B reliability-bootstrap diagnostics.",
  align = c(
    "l",
    "r"
  )
)
```

All 2,000 bootstrap resamples completed successfully.

No resampled polychoric matrix required smoothing, and no resample produced a computational warning. The minimum raw eigenvalue observed across the resampled matrices was 0.171.

# Stage 2B validation checks

The following checks were completed:

- the Stage 2B complete-PHQ-9 sample contains 8,276 participants;
- the Stage 2B item matrix contains nine items;
- all item values fall within the valid range from 0 to 3;
- the Stage 2B item matrix contains no missing values;
- each item-response distribution sums to 8,276 participants;
- each item-response distribution sums to 100% within numerical tolerance;
- no PHQ-9 item has an empty response category;
- only `DPQ090` has response categories below the 1% screening threshold;
- manually calculated corrected item-total correlations match `psych::alpha()` `r.drop` values within numerical precision;
- the maximum item-total validation difference was approximately `6.13e-14`;
- the raw polychoric matrix is finite, symmetric and positive definite;
- the raw polychoric matrix has a minimum eigenvalue of 0.204;
- the polychoric matrix did not require smoothing;
- the manually calculated ordinal alpha matches the standardised alpha from `psych::alpha()` exactly;
- the two algebraically equivalent omega-total calculations agree within approximately `1.11e-16`;
- all nine uniqueness estimates used for omega total were finite, non-negative and no greater than 1, with values ranging from 0.236 to 0.525;
- removing any individual item lowers both ordinal alpha and omega total;
- all 2,000 final bootstrap resamples completed successfully;
- no final bootstrap replicate required matrix smoothing;
- no final bootstrap replicate produced a computational warning.

The final bootstrap object is saved locally as:

```text
data/processed/phq9_reliability_bootstrap_stage2b.rds
```

This processed object is excluded from version control.

# Analyses not included in Stage 2B

Stage 2B is limited to reliability and item analysis.

The following analyses remain outside the present stage:

- exploratory factor analysis;
- confirmatory factor analysis;
- evaluation of one-factor model fit;
- comparison of alternative factor structures;
- measurement-invariance testing;
- survey-weighted latent-variable modelling;
- functional-difficulty regression modelling;
- diagnostic-accuracy analysis;
- diagnostic interpretation;
- treatment-need claims;
- substantive clinical interpretation.

A single-factor congeneric model was used only to obtain the uniqueness estimates required for omega total. Its factor loadings and model fit were not interpreted as evidence about the dimensional structure of the PHQ-9.

The Stage 2B findings provide evidence about internal consistency and item relationships in the complete-PHQ-9 analytic sample. They do not establish validity, unidimensionality, diagnostic accuracy, psychiatric diagnosis or treatment need.