---
title: "Psychometric Evaluation of the PHQ-9 in U.S. Adults Using NHANES 2017–March 2020"
subtitle: "Reliability, Dimensionality, Functional Difficulty, and Clinical Interpretation"
author: "Shane Hodson"
date: "15 July 2026"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(readr)
library(here)
library(knitr)

display_audit_table <- function(file_name, caption) {
  
  file_path <- here("tables", file_name)
  
  if (file.exists(file_path)) {
    
    read_csv(
      file_path,
      show_col_types = FALSE
    ) |>
      kable(
        caption = caption,
        digits = 2
      )
    
  } else {
    
    cat(
      paste0(
        "**Output not yet generated:** `",
        file_name,
        "`"
      )
    )
  }
}
```

# Project status

This report documents the repository setup and initial NHANES data-audit stage.

Reliability analysis, factor analysis and substantive clinical interpretation have not yet begun.

# Dataset

The project uses:

- `P_DPQ.xpt`: Mental Health—Depression Screener
- `P_DEMO.xpt`: Demographic and survey-design variables

The two files are linked using the NHANES respondent identifier `SEQN`.

# Source-file audit

```{r}
display_audit_table(
  "data_file_audit.csv",
  "Initial NHANES source-file audit"
)
```

# SEQN linkage audit

```{r}
display_audit_table(
  "seqn_linkage_audit.csv",
  "Linkage audit using SEQN"
)
```

All depression-screener records were successfully matched to the demographic file.

Adults appearing in the demographic file but not in the depression-screener file are retained in the sample-flow documentation but are not added artificially to the depression analysis sample.

# PHQ-9 variable dictionary

```{r}
display_audit_table(
  "phq9_variable_dictionary.csv",
  "PHQ-9 and functional-difficulty variable dictionary"
)
```

# Response-code audit

```{r}
display_audit_table(
  "response_code_audit.csv",
  "Observed questionnaire response codes"
)
```

The audit distinguishes valid responses from refusal, "don't know" and system-missing values. No unexpected response codes were identified.

# Missingness audit

```{r}
display_audit_table(
  "missingness_audit.csv",
  "Item-level missingness audit"
)
```

# DPQ100 pattern audit

```{r}
display_audit_table(
  "dpq100_skip_audit.csv",
  "DPQ100 status by PHQ-9 response pattern"
)
```

All participants with valid zero responses to all nine PHQ-9 items had a system-missing `DPQ100` response.

Among participants endorsing at least one PHQ-9 symptom, nearly all had a valid `DPQ100` response.

This pattern is consistent with structural questionnaire routing rather than ordinary item non-response. No recoding decision is implemented at this stage.

# Initial sample flow

```{r}
display_audit_table(
  "sample_flow.csv",
  "Initial analytic sample flow"
)
```

# Survey-design variable audit

```{r}
display_audit_table(
  "survey_design_audit.csv",
  "NHANES survey-design variable audit"
)
```

The combined-cycle examination weight, masked variance stratum and masked variance primary sampling unit variables were present for all linked depression-screener records.

# Cleaned questionnaire-code check

```{r}
display_audit_table(
  "clean_code_check.csv",
  "Questionnaire values after recoding non-substantive responses"
)
```

Refusal, "don't know" and system-missing responses were recoded as `NA`.

After cleaning, each PHQ-9 symptom item contained only valid values from 0 to 3 or missing values. No unexpected codes remained.

Structurally missing `DPQ100` responses have not been recoded as a functional-difficulty category.

# Data-audit decisions

The current audit establishes that:

- `P_DPQ` contains 8,965 unique respondents;
- all `P_DPQ` respondents link successfully to `P_DEMO`;
- 8,276 respondents have valid responses to all nine PHQ-9 items;
- 5,517 respondents have both a complete PHQ-9 and a valid `DPQ100` response;
- non-substantive values must be recoded as missing before later analysis;
- `DPQ100` system missingness is strongly associated with reporting zero on all nine symptom items;
- all required survey-design variables are available.

# Scope boundary

No reliability coefficients, polychoric correlations, factor models or clinical conclusions are reported at this stage.
