---
title: "Psychometric Evaluation of the PHQ-9 in U.S. Adults Using NHANES 2017–March 2020"
subtitle: "Reliability, Dimensionality, Functional Difficulty, and Clinical Interpretation"
author: "Shane Hodson"
date: "22 July 2026"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
    fig-cap-location: bottom
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(dplyr)
library(readr)
library(here)
library(knitr)

display_audit_table <- function(file_name, caption, digits = 2) {

  file_path <- here("tables", file_name)

  if (file.exists(file_path)) {

    read_csv(
      file_path,
      show_col_types = FALSE
    ) |>
      kable(
        caption = caption,
        digits = digits
      )

  } else {

    cat(
      paste0(
        "**Output not yet generated:** `",
        file_name,
        "`"
      )
    )
  }
}

read_stage2a_table <- function(file_name) {

  file_path <- here("tables", file_name)

  if (!file.exists(file_path)) {
    stop(
      paste0(
        "Required Stage 2A table not found: ",
        file_path
      )
    )
  }

  read_csv(
    file_path,
    show_col_types = FALSE
  )
}

format_percent_ci <- function(
  estimate,
  lower,
  upper,
  digits = 2
) {

  paste0(
    formatC(
      estimate,
      format = "f",
      digits = digits
    ),
    "% (",
    formatC(
      lower,
      format = "f",
      digits = digits
    ),
    "–",
    formatC(
      upper,
      format = "f",
      digits = digits
    ),
    "%)"
  )
}

display_figure <- function(file_name) {

  figure_path <- here(
    "figures",
    file_name
  )

  if (file.exists(figure_path)) {

    include_graphics(
      figure_path
    )

  } else {

    cat(
      paste0(
        "**Figure not yet generated:** `",
        file_name,
        "`"
      )
    )
  }
}
```

# Project status

This report documents the completed data audit and the Stage 2A weighted descriptive analyses.

Stage 2A includes:

- construction and validation of the NHANES survey-design object;
- weighted PHQ-9 item-response proportions;
- weighted PHQ-9 total-score descriptives;
- weighted symptom-severity band estimates;
- weighted `DPQ100` descriptives among routed respondents;
- an extended functional-difficulty distribution that retains structural skips as a separate category;
- exported descriptive tables and figures.

Reliability, omega, polychoric correlations, factor analysis, functional-difficulty modelling and clinical interpretation have not begun.

# Dataset

The project uses the combined NHANES 2017–March 2020 pre-pandemic release.

The primary files are:

- `P_DPQ.xpt`: Mental Health—Depression Screener
- `P_DEMO.xpt`: demographic and survey-design variables

The files are linked using the NHANES respondent identifier `SEQN`.

Raw NHANES files are downloaded programmatically and are not redistributed through the repository.

# Stage 1: Data audit

## Source-file audit

```{r}
display_audit_table(
  "data_file_audit.csv",
  "Initial NHANES source-file audit"
)
```

## SEQN linkage audit

```{r}
display_audit_table(
  "seqn_linkage_audit.csv",
  "Linkage audit using SEQN"
)
```

All 8,965 depression-screener records were successfully matched to the demographic file.

Adults appearing in the demographic file but not in the depression-screener file are retained in the sample-flow documentation but are not added to the depression analysis sample.

## PHQ-9 variable dictionary

```{r}
phq9_dictionary_table <- read_stage2a_table(
  "phq9_variable_dictionary.csv"
) |>
  transmute(
    Variable = variable,
    `Item number` = item_number,
    `Symptom content` = content,
    Role = role,
    `Included in PHQ-9 total` = if_else(
      variable == "DPQ100",
      "No",
      "Yes"
    )
  )

kable(
  phq9_dictionary_table,
  caption = "PHQ-9 and functional-difficulty variable dictionary",
  align = c(
    "l",
    "r",
    "l",
    "l",
    "l"
  )
)
```

## Response-code audit

```{r}
display_audit_table(
  "response_code_audit.csv",
  "Observed questionnaire response codes"
)
```

The audit distinguishes valid responses from refusal, “don’t know” and system-missing values.

No unexpected questionnaire response codes were identified.

## Missingness audit

```{r}
display_audit_table(
  "missingness_audit.csv",
  "Item-level missingness audit"
)
```

## DPQ100 routing audit

```{r}
display_audit_table(
  "dpq100_skip_audit.csv",
  "DPQ100 status by PHQ-9 response pattern"
)
```

All participants with valid zero responses to all nine PHQ-9 items had a system-missing `DPQ100` response.

Among participants endorsing at least one PHQ-9 symptom, nearly all had a valid `DPQ100` response.

This pattern is consistent with structural questionnaire routing among participants reporting no PHQ-9 symptoms.

Structurally skipped `DPQ100` responses are not recoded as “not difficult at all.”

## Initial sample flow

```{r}
display_audit_table(
  "sample_flow.csv",
  "Initial analytic sample flow"
)
```

## Survey-design variable audit

```{r}
display_audit_table(
  "survey_design_audit.csv",
  "NHANES survey-design variable audit"
)
```

The combined-cycle examination weight, masked variance stratum and masked variance primary sampling unit variables were present for all 8,965 linked depression-screener records.

## Cleaned questionnaire-code check

```{r}
display_audit_table(
  "clean_code_check.csv",
  "Questionnaire values after recoding non-substantive responses"
)
```

Refusal, “don’t know” and system-missing responses were treated as missing values.

After cleaning, each PHQ-9 symptom item contained only valid values from 0 to 3 or missing values.

# Stage 2A: Weighted descriptive analyses

## Analytic samples

The linked depression-screener dataset contains 8,965 adults.

The analysis-specific samples are:

- **8,276 participants** with valid responses to all nine PHQ-9 items;
- **5,522 participants** with complete PHQ-9 data and at least one endorsed symptom;
- **5,517 participants** with complete PHQ-9 data, at least one endorsed symptom and valid `DPQ100`;
- **2,754 participants** with zero on all nine PHQ-9 items and structurally skipped `DPQ100`;
- **5 participants** with complete PHQ-9 data, at least one endorsed symptom and missing `DPQ100`.

The complete-PHQ-9 sample is used for total-score and symptom-severity descriptives.

Item-response estimates use all available valid responses for each individual item.

The primary `DPQ100` distribution is restricted to routed respondents with valid responses.

## Survey design

The survey-design object was created from the full linked depression-screener dataset using:

- examination weight: `WTMECPRP`;
- masked variance stratum: `SDMVSTRA`;
- masked variance primary sampling unit: `SDMVPSU`;
- nested PSU identifiers within strata.

The design contains 49 nested PSUs across 24 strata.

No observed stratum contained fewer than two PSUs.

Analysis-specific samples were defined as survey domains after the full survey-design object had been created.

Weighted means and item-response proportions were estimated using `svymean()`. Item-response confidence intervals were calculated from the design-based standard errors.

Weighted variances were estimated using `svyvar()`.

Proportions for individual total scores, severity bands and `DPQ100` categories were estimated using `svyciprop()` with logit-transformed 95% confidence intervals.

## Weighted PHQ-9 item-response proportions

Item-specific valid sample sizes ranged from 8,295 to 8,299.

The weighted response proportions sum to 100% within rounding error for each item.

```{r}
item_response_table <- read_stage2a_table(
  "phq9_item_response_proportions.csv"
) |>
  transmute(
    Item = item,
    `Symptom content` = symptom_content,
    Response = response_label,
    `Unweighted n` = unweighted_n_response,
    `Weighted estimate (95% CI)` = format_percent_ci(
      weighted_percent,
      ci_lower_percent,
      ci_upper_percent
    )
  )

kable(
  item_response_table,
  caption = paste(
    "Weighted PHQ-9 item-response proportions.",
    "Item-specific estimates use all available valid responses."
  ),
  align = c(
    "l",
    "l",
    "l",
    "r",
    "r"
  )
)
```

```{r}
#| fig-alt: "Stacked horizontal bar chart showing weighted response proportions for each of the nine PHQ-9 items."
#| out-width: "100%"

display_figure(
  "phq9_item_response_proportions.png"
)
```

## Weighted PHQ-9 total-score distribution

The complete-PHQ-9 sample contains 8,276 participants.

Observed PHQ-9 total scores range from 0 to 27.

```{r}
total_summary_raw <- read_stage2a_table(
  "phq9_total_summary.csv"
)

total_summary_table <- tibble(
  Measure = c(
    "Unweighted sample size",
    "Weighted mean",
    "Standard error of weighted mean",
    "95% confidence interval for weighted mean",
    "Weighted standard deviation",
    "Observed minimum",
    "Observed maximum"
  ),
  Estimate = c(
    format(
      total_summary_raw$unweighted_n,
      big.mark = ","
    ),
    formatC(
      total_summary_raw$weighted_mean,
      format = "f",
      digits = 2
    ),
    formatC(
      total_summary_raw$standard_error,
      format = "f",
      digits = 2
    ),
    paste0(
      formatC(
        total_summary_raw$ci_lower,
        format = "f",
        digits = 2
      ),
      "–",
      formatC(
        total_summary_raw$ci_upper,
        format = "f",
        digits = 2
      )
    ),
    formatC(
      total_summary_raw$weighted_standard_deviation,
      format = "f",
      digits = 2
    ),
    total_summary_raw$observed_minimum,
    total_summary_raw$observed_maximum
  )
)

kable(
  total_summary_table,
  caption = "Weighted PHQ-9 total-score summary.",
  align = c(
    "l",
    "r"
  )
)
```

The weighted mean PHQ-9 total score was 3.18, with a 95% confidence interval from 3.06 to 3.31.

The weighted standard deviation was 4.09.

```{r}
total_distribution_table <- read_stage2a_table(
  "phq9_total_distribution.csv"
) |>
  transmute(
    `PHQ-9 total` = phq9_total,
    `Unweighted n` = unweighted_n,
    `Weighted estimate (95% CI)` = format_percent_ci(
      weighted_percent,
      ci_lower_percent,
      ci_upper_percent
    )
  )

kable(
  total_distribution_table,
  caption = paste(
    "Weighted distribution of PHQ-9 total scores",
    "in the complete-PHQ-9 sample."
  ),
  align = c(
    "r",
    "r",
    "r"
  )
)
```

```{r}
#| fig-alt: "Bar chart showing the weighted distribution of PHQ-9 total scores from 0 to 27."
#| out-width: "100%"

display_figure(
  "phq9_total_score_distribution.png"
)
```

## Weighted symptom-severity bands

The conventional score bands used in this descriptive analysis are:

- minimal: 0–4;
- mild: 5–9;
- moderate: 10–14;
- moderately severe: 15–19;
- severe: 20–27.

These categories describe score ranges and are not treated as psychiatric diagnoses.

```{r}
severity_table <- read_stage2a_table(
  "phq9_severity_distribution.csv"
) |>
  transmute(
    `Severity band` = severity_band,
    `Unweighted n` = unweighted_n,
    `Weighted estimate (95% CI)` = format_percent_ci(
      weighted_percent,
      ci_lower_percent,
      ci_upper_percent
    )
  )

kable(
  severity_table,
  caption = paste(
    "Weighted PHQ-9 symptom-severity bands",
    "in the complete-PHQ-9 sample."
  ),
  align = c(
    "l",
    "r",
    "r"
  )
)
```

The weighted distribution was 74.92% minimal, 16.54% mild, 5.95% moderate, 1.81% moderately severe and 0.78% severe.

```{r}
#| fig-alt: "Horizontal bar chart showing weighted PHQ-9 symptom-severity band percentages with 95% confidence intervals."
#| out-width: "100%"

display_figure(
  "phq9_severity_distribution.png"
)
```

## DPQ100 among routed respondents

The primary functional-difficulty descriptive sample contains 5,517 participants who:

- had complete PHQ-9 data;
- endorsed at least one PHQ-9 symptom;
- provided a valid `DPQ100` response.

This section reports the weighted response distribution only.

It does not model the relationship between PHQ-9 scores and functional difficulty.

```{r}
dpq100_routed_table <- read_stage2a_table(
  "dpq100_routed_distribution.csv"
) |>
  transmute(
    Response = response_label,
    `Unweighted n` = unweighted_n,
    `Weighted estimate (95% CI)` = format_percent_ci(
      weighted_percent,
      ci_lower_percent,
      ci_upper_percent
    )
  )

kable(
  dpq100_routed_table,
  caption = paste(
    "Weighted DPQ100 distribution among routed respondents",
    "with valid functional-difficulty data."
  ),
  align = c(
    "l",
    "r",
    "r"
  )
)
```

Among routed respondents, 74.64% reported “not difficult at all,” 20.79% reported “somewhat difficult,” 3.40% reported “very difficult,” and 1.17% reported “extremely difficult.”

```{r}
#| fig-alt: "Horizontal bar chart showing weighted DPQ100 response percentages among routed respondents with 95% confidence intervals."
#| out-width: "100%"

display_figure(
  "dpq100_routed_distribution.png"
)
```

## Extended functional-difficulty categories

The extended table uses the complete-PHQ-9 sample of 8,276 participants.

Participants with zero on all nine PHQ-9 items are retained in a separate category:

> No PHQ-9 symptoms / DPQ100 structurally skipped

The five symptom-positive participants without a valid `DPQ100` response are also shown separately.

Neither group is assigned to an observed functional-difficulty response category.

```{r}
dpq100_extended_table <- read_stage2a_table(
  "dpq100_extended_distribution.csv"
) |>
  transmute(
    Category = category,
    `Unweighted n` = unweighted_n,
    `Weighted estimate (95% CI)` = format_percent_ci(
      weighted_percent,
      ci_lower_percent,
      ci_upper_percent
    )
  )

kable(
  dpq100_extended_table,
  caption = paste(
    "Extended weighted functional-difficulty distribution",
    "in the complete-PHQ-9 sample."
  ),
  align = c(
    "l",
    "r",
    "r"
  )
)
```

```{r}
#| fig-alt: "Horizontal bar chart showing extended weighted functional-difficulty categories, including structural skips and missing DPQ100 responses."
#| out-width: "100%"

display_figure(
  "dpq100_extended_distribution.png"
)
```

# Stage 2A validation checks

The following checks were completed:

- the linked depression-screener sample contains 8,965 records;
- the complete-PHQ-9 sample contains 8,276 records;
- the symptom-positive complete-PHQ-9 sample contains 5,522 records;
- the routed valid-`DPQ100` sample contains 5,517 records;
- the structural-skip group contains 2,754 records;
- the weighted response proportions sum to 100% within rounding error for each PHQ-9 item;
- the score-distribution counts sum to 8,276;
- the weighted score-distribution proportions sum to 100% within rounding error;
- the severity-band counts sum to 8,276;
- the weighted severity-band proportions sum to 100% within rounding error;
- the routed `DPQ100` counts sum to 5,517;
- the weighted routed-`DPQ100` proportions sum to 100% within rounding error;
- the extended functional-difficulty counts sum to 8,276;
- the weighted extended-category proportions sum to 100% within rounding error;
- no symptom-free participant had a valid `DPQ100` response;
- five symptom-positive participants had missing `DPQ100` responses and remain classified as missing.

The completed survey-design object is saved locally as:

```text
data/processed/nhanes_design_stage2a.rds
```

The Stage 2A working dataset is saved locally as:

```text
data/processed/phq9_stage2a_data.rds
```

These processed files are excluded from version control.

# Analyses not included in Stage 2A

Stage 2A is limited to weighted descriptive analysis and sample reporting.

The following analyses are outside the present stage:

- ordinal coefficient alpha;
- McDonald’s omega;
- corrected item-total correlations;
- polychoric correlations;
- factor analysis;
- functional-difficulty regression modelling;
- diagnostic interpretation;
- treatment-need claims.

No conclusions about validity, unidimensionality, psychiatric diagnosis or treatment need are drawn from the Stage 2A descriptive results.